---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
```{r tweak locale, echo=FALSE, results='hide'}
Sys.setlocale(category = "LC_ALL", locale = "English_United States.1252")
```

## Loading and preprocessing the data
1. Load the data using read.csv(), save the data to act

```{r read data}
act <- read.csv("activity.csv", stringsAsFactors = FALSE)
```
2. Process the data, sum steps by day

```{r sum steps by day}
sum_by_day <- tapply(act$steps, act$date, sum, na.rm = TRUE)
```

## What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day

```{r histogram of sum by day, fig.height=4}
hist(sum_by_day,
     main = "Histogram of the total number of steps taken each day",
     xlab = "Sum of daily steps")
```

2. Calculate and report the mean and median total number of steps taken per day

```{r mean and median}
mean_by_day <- tapply(act$steps, act$date, mean, na.rm = TRUE)
median_by_day <- tapply(act$steps, act$date, median, na.rm = TRUE)
rep <- cbind.data.frame(names(mean_by_day),
                        as.numeric(mean_by_day),
                        as.numeric(median_by_day))
names(rep) <- c("date","mean","median")
rep
```

## What is the average daily activity pattern?
1. Make a time series plot(i.e. type = "l") of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis)

```{r average daily steps,fig.height=4}
mean_by_interval = tapply(act$steps, act$interval, mean, na.rm = TRUE)
plot(
    mean_by_interval,
    type = "l",
    main = "Average number of steps by interval",
    ylab = "average number of steps",
    xlab = "5-minute interval"
    )
```

2. Which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps?

```{r find max}
which(mean_by_interval == max(mean_by_interval))
```
It's the 8:35 interval containing the maximum number of steps.

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r how many missing values}
sum(is.na(act$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset.
Here, the strategy is inputting NA using the mean for that 5-minute interval.

```{r function for imputting mussing values}
# define the funtion to input missing values
imputNA <- function(rowData) {
    if (is.na(rowData[1]))
        return(mean_by_interval[[paste0(as.numeric(rowData[3]))]])
    else
        return(rowData[1])
    }
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r create new dataset}
# new dataset, NA is replaced by interval mean
act2 = cbind(apply(act,1,imputNA), act[2:3])
# fix column names
names(act2) = c(names(act))
# convert steps to numeric
act2$steps = as.numeric(act2$steps)
```

4. Make a histogram of the total number of steps taken each day and Calculate
and report the mean and median total number of steps taken per day.

```{r new dataset, fig.height=4}
sum_by_day_new <- tapply(act2$steps, act2$date, sum)
hist(sum_by_day_new,
     main = "Histogram of the total number of steps taken each day(new dataset)",
     xlab = "Sum of daily steps")
mean_by_day_new = tapply(act2$steps, act2$date, mean)
median_by_day_new = tapply(act2$steps, act2$date, median)
rep2 = cbind.data.frame(
    names(mean_by_day_new),
    as.numeric(mean_by_day_new),
    as.numeric(median_by_day_new)
    )
names(rep2) <- c("date","mean","median")
rep2
```

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels -- "weekday" and
"weekend" indicating whether a given date is a weekday or weekend day

```{r new factor variable}
# define a function to process weekday and weekend
weekdayorweekend <- function (x) {
    w = weekdays(as.Date.character(x), abbreviate = TRUE)
    if (w == "Sun" | w == "Sat")
        result = "weekend"
    else
        result = "weekday"
    return(result)
    }
# add new variable wday to act2
act2 = cbind(act2,
             as.vector(sapply(act$date,
             weekdayorweekend,
             simplify = TRUE)))
names(act2) <- c(names(act2)[1:3], "wday")
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the
5-minute interval (x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis).

```{r weekdayvsweekend, fig.height=4}
library(plyr)
library(lattice)
ddata = ddply(act2, c("interval","wday"), summarise, mean = mean(steps, na.rm =
                                                                     TRUE))
xyplot(
    mean ~ interval | wday, layout = c(1,2), data = ddata, type = "l",
    xlab = "Interval",
    ylab = "Number of steps"
    )
```
